---
title: "Final Project"
author: "Lala"
execute:
  echo: true
  warning: false
format: 
  html:
    embed-resources: true
--- 

# Measles cases across the world

## use closeread

Data from https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-06-24/readme.md 

the questions posed om this repo were:

- How have global measles cases changed over time?

- Which regions or countries consistently report the highest measles burden?

- Are there seasonal patterns in measles outbreaks across different regions?

- Does the ratio of laboratory-confirmed cases to total cases reveal differences in healthcare capacity across countries?

```{r}
#| warning: false
library(tidyverse)
library(readr)

measles <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')

View(measles)

unique(measles$year)
```


## using incidence rates (cases per million people) rather than actual cases reported to avoid bigger-population countries having more cases just by virtue of being large??? 


# talk to Dr. Vilakati in the public health department to get more insight on what lens to look at the data from

How have global measles cases changed over time?

- line plot to show trends in total measles cases over the years - interpretation - growing or falling?

using actual measles cases reported - not incidence rates?

```{r}
#making a data frame with the total global measles cases reports for each year 

annual_measles <- measles |> group_by(year) |>
  summarise(total_cases = sum(measles_total))
  
ggplot(data = annual_measles, aes(x = year, y = total_cases)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Trend of measles cases throughout the world for the past 14 years",
       x = "Years",
       y = "Total global cases")
```


- were there years when there were global outbreaks?

linear regression model to visualize the trends 



# what is the spatial distribution of measles cases across the world?

- choropleth map showing the spatial distribution of measles cases across the world

- make it interactive with leaflet

# Which regions or countries consistently report the highest measles cases per (eg 10 000 of the population)?

-- comparison of Boxplots of incidence rates by region  -- continent??

# Are there seasonal patterns in measles outbreaks across different regions?

- using monthly data  -- heatmap (month vs year)

# Does the ratio of laboratory-confirmed cases to total cases reveal differences in healthcare capacity across countries?

- comparing proportions of lab-confirmed vs clinical vs suspected cases 

-- Stacked bar chart??

# Data quality classification:  Which countries have strong lab-confirmation systems?

high_quality = ifelse(lab_confirmed / total > 0.5, 1, 0)

-- maybe join the data set with economic data to make a logistic regression model to predict data quality classification based on eg GDP, population size, region, ...

-- but what would data quality add to the story telling??? 

## Modeling mealses cases using predictors that are economic, and maybe rubella cases as well?/??

Future work:

vaccination data
policy analysis





- 