---
title: "README"
format: gfm
editor: visual
---

### The global case of Measles

Measles is a highly contagious viral disease that remains a public health concern despite the availability of an effective vaccine. This project examines global trends in measles incidence from 2012 to 2024 using country-level data on reported cases, incidence rates, and case classifications. The analysis explores how measles incidence has changed over time, how it varies geographically, and which countries experience the highest disease burden.
Using spatial and time-series visualizations, the project identifies patterns in measles distribution across regions and highlights countries with high incidence rates, particularly in parts of Central Asia and Eastern Europe for 2024. 

In addition, the analysis considers the potential role of economic stability in the incidence rate of measles through modeling the influence of GDP per Capita on incidence rates.

Overall, the findings show that measles incidence is unevenly distributed across the world and influenced by in part by economic factors.

## Questions explored

1. How have measles cases changed globally over time?

2. Geographic distribution of measles cases across countries and regions

3. Which countries and regions experienced the highest measles incidence rates in 2024?

4. How do measles cases differ between Lesotho and USA?

### The data 
```{r}
#| echo: false
#| warning: false
library(tidyverse)
library(readr)

measles <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
measles_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
# the data for the year 2025 is incomplete for all countries. 2025 records in measles_month contain 4 to 6 months' data 
# so, I will filter out the year 2025 to eliminate these incomplete records for the sake of consistency throughout the years
measles <- measles |> select(1,2,3, 4, 5, 8, 9, 10, 11, 12) |> 
  filter(year != 2025) 

#preview of data 
measles_preview <- measles |> group_by(region) |>
  slice(1)

library(knitr)
measles_preview |> kable()
```

### Global trend in total cases of measles over the 13 years
```{r}
#| echo: false
#Line plot for global trend in total cases of measles over the years

annual_measles <- measles |> group_by(year) |>
  summarise(`total measles cases in millions` = sum(measles_total)/100000)
  
ggplot(data = annual_measles, aes(x = year, y = `total measles cases in millions`)) +
  geom_line(color = "darkblue") +
  theme_minimal(base_size = 24) +
  labs(x = "Years",
       y = "Total global measles cases in millions")
```
