---
title: "The Case Of Measles"
author: "Lala"
execute:
  echo: true
  warning: false
format: 
  html:
    embed-resources: true
--- 

# Abstract
Measles is a highly contagious viral disease that remains a public health concern despite the availability of an effective vaccine. This project examines global trends in measles incidence from 2012 to 2024 using country-level data on reported cases, incidence rates, and case classifications. The analysis explores how measles incidence has changed over time, how it varies geographically, and which countries experience the highest disease burden.
Using spatial and time-series visualizations, the project identifies patterns in measles distribution across regions and highlights countries with high incidence rates, particularly in parts of Central Asia and Eastern Europe for 2024. 

?? In addition, the analysis considers the potential role of surveillance intensity in shaping reported incidence and compares trends between countries with different socioeconomic and health system contexts.
Overall, the findings show that measles incidence is unevenly distributed across the world and influenced by both epidemiological and reporting factors. The project emphasizes the importance of vaccination coverage and reliable surveillance systems in understanding and controlling measles outbreaks globally.

# Introduction

According to the CDC, measles is a highly contagious viral disease that spreads through the air and can lead to serious complications, especially in young children below the age of 5. Symptoms include a rash, fever, coughing and a runny nose. Although it is preventable through vaccination, it remains a concern in many parts of the world. [[cite https://www.cdc.gov/measles/about/index.html]]

Growing up in Lesotho, I used to hear about measles a lot, either when a classmate had contracted it and therefore was taking some time off from school, or when my mom would take me to doctors' appointments and one of the first things they would ask her would be whether I had gotten a measles/rubella vaccination. Since then I hadn't heard or read much about measles until I came across this dataset. Finding this dataset sparked my curiosity about whether I had stopped hearing about measles because it wasn't as rampant a disease anymore or if my change of location from Lesotho to Eswatini and now the U.S. within the past seven years had anything to do with it.

The project explores the trends in reported measles cases across the world over a period of 13 years from 2012 to 2024. The exploration with a broad, global perspective using spatial visualizations of the cases as well as an analysis of trends of measles cases over time then, zooms in on regions of interest to tell a more specific story. There are some micro-level comparisons of the distribution of measles between geographic regions and an assessment of the role played by economic factors on the reporting and confirmation of measles cases.

The data for this project was sourced from a tidyTuesday data GitHub repository titled (["Measles cases across the world"](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-06-24/readme.md)), which includes country-level data on measles incidence reporting, population, and case classification metrics.

To satisfy my curiosity, here are the questions I will be exploring:

1. How have measles cases changed globally over time?

2. Geographic distribution of measles cases across countries and regions

3. Which countries and regions experienced the highest measles incidence rates in 2024?

4. How do measles cases differ between Lesotho and USA?


## Data

```{r}
#| echo: false
#| warning: false
library(tidyverse)
library(readr)

measles <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')

measles_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')

# the data for the year 2025 is incomplete for all countries. 2025 records in measles_month contain 4 to 6 months' data 
# so, I will filter out the year 2025 to eliminate these incomplete records for the sake of consistency throughout the years

measles <- measles |> select(1,2,3, 4, 5, 8, 9, 10, 11, 12) |> 
  filter(year != 2025) 

#preview of data 
measles_preview <- measles |> group_by(region) |>
  slice(1)

library(knitr)
measles_preview |> kable()
```

The data for the year 2025 is incomplete for all countries.The data set containing monthly data only consists of 4 to 6 months' data for 2025 for all countries, implying that at the time this data set was published, the data collection process for 2025 was not yet completed. So, I have filtered out the year 2025 to eliminate these incomplete records for the sake of consistency throughout the years.

To explore the questions of interest mentioned above, the following sections use visualizations such as line plots and choropleth maps to display trends in measles cases over time and across the world.

## The global trend in total cases of measles over the 13 years from 2012 to 2024 (inclusive)

```{r}
#| echo: false
#Line plot for global trend in total cases of measles over the years

annual_measles <- measles |> group_by(year) |>
  summarise(`total measles cases in millions` = sum(measles_total)/100000)
  
ggplot(data = annual_measles, aes(x = year, y = `total measles cases in millions`)) +
  geom_line(color = "purple") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2012, 2024, by = 2)) +
  scale_y_continuous(limits = c(0,6)) +   #referenced
  labs(title = "Trend of measles cases throughout the world for the past 13 years",
       x = "Years",
       y = "Total global measles cases in millions")
```

The line plot above shows a fluctuating trend in total measles cases reported across the world between 2012 and 2024. There are three major peaks around 2014, 2019 and 2024. These suggest that there were some outbreaks or surges in total reported cases in the world in those periods. There are two major dips int he trend between these three peaks, corresponding to 2016-2017 and 2020-2021. Interestingly, in 2020-2021, measles cases reported were the lowest in that 13-year period. Since 2020-2021 is the time notoriously known for the Covid-19 pandemic when people were "forced" to isolate, this may have decreased the rate of infections of measles as people did not come into contact with infected individuals. However, there is also a possibility that the low records of measles cases in that period just reflect the fact that medical resources and reporting was aimed more towards tracking the Covid-19 pandemic and measles reporting was not prioritized as much as in other years.

As of 2024, the total cases of measles across the world were on a sharp rise. This contradicts human expectations that as the world at large advances technologically and medically, the spread of preventable diseases like measles would decrease.

This increasing trend in measles cases begs the question: which countries contribute the most to this growth? To answer this, below is a choropleth map that shows the global distribution of total measles cases.

```{r}
measles_2024 <- measles |> filter(year == 2024)


library(maps)
world <- map_data("world")

measles_2024 <- measles_2024 |> mutate(country = if_else(country == "United States of America",
                                   true = "USA",
                                   false = country))

measles_2024 <- measles_2024 |> mutate(country = if_else(country == "Russian Federation",
                                   true = "Russia",
                                   false = country))

measles_2024 <- measles_2024 |> mutate(country = if_else(country == "Eswatini",
                                   true = "Swaziland",
                                   false = country))


world_measles <- measles_2024 |> left_join(world, by = join_by(country == region))

plot <- ggplot(data = world_measles, aes(x = long, y = lat, group = group, text = paste0("Country: ", country, "\n",
    "Total cases: ",  measles_total)))  + #referenced
  geom_polygon(colour = "black", aes(fill = measles_total)) +
  coord_map(projection = "gilbert") +
  theme_void() +
  scale_fill_viridis_c() +
  xlim(c(-180, 180)) + 
  labs(fill = "Total cases reported", 
       title = "Global distribution of measles cases" )

library(plotly)

ggplotly(plot, tooltip = "text")
```

Looking at the map above, most of the world's countries had less than 10,000 reported cases in 2024. However a select few reached 20,000 cases and above. As you hover over the map, you can see the names of each country and the total number of cases it reported in 2024. 

The barplot below shows the top 10 countries with the highest total number of measles cases reported in 2024 for easier comparison.

```{r}
library(forcats)

top10_cases <- measles_2024 |> arrange(desc(measles_total)) |>
  slice(1:10) |>
  mutate(country = fct_reorder(country,
    measles_total))

top10_plot <- ggplot(data = top10_cases, aes(x = country, y = measles_total, cases = measles_total)) +
  geom_col(aes(fill = region)) +
  theme_minimal() +
  labs(y = "Total reported cases") +
  coord_flip() +
  scale_fill_viridis_d()

ggplotly(top10_plot, tooltip = "cases")

measles_2024 |> filter(country == "Namibia") 
```

From the plot above, we can observe that in 2024, Iraq reported the highest number of cases in the world with 32400 cases. 

The World Health Organization (WHO) groups its member states into these 6 regions: Africa (AFRO), the Americas (AMRO), South-East Asia (SEARO), Europe (EURO), Eastern Mediterranean (EMRO), and the Western Pacific (WPRO). According to WHO, these regions do not strictly follow continental boundaries. As seen on the plot above, Kyrgyzstan, which is in Central Asia is categorized under the EURO region. 

From the barplot, half of the top ten countries with the highest reported cases of measles in 2024 were in the EURO region, 3 were in the EMRO region and there was only 1 country for the SEARO and AFRO regions each.

Although total reported cases provide an indication of the spread of the disease throughout a country, they are not necessarily a good metric for comparing the burden of the disease among various countries with different populations. A total count of 32400 in Iraq in 2024 where the population was around 46 million and a total count of 74 in Namibia where the population was around 3 million in 2024 do not give a fair comparison of the prevalence of the disease and level of risk for the people in that country. 

So, to assess this risk and prevalence, the plots below use incidence rate per 1 million of the population to compare measles reports between countries for the year 2024.

Incidence rate is a measure of new cases of a disease within a particular population over a set period of time. In this case, incidence rate measures the number of new case per 1 million of aeach country's population for a given year.

```{r}
#| echo: false

map_inc <- ggplot(data = world_measles, aes(x = long, y = lat, group = group, text = paste0("Country: ", country, "\n",
    "Incidence: ", measles_incidence_rate_per_1000000_total_population))) +
  geom_polygon(colour = "black", aes(fill = measles_incidence_rate_per_1000000_total_population)) +
  coord_map(projection = "gilbert") +
  theme_void() +
  scale_fill_viridis_c() +
  xlim(c(-180, 180)) + 
  labs(fill = "Incidence rate per million")

ggplotly(map_inc, tooltip = "text")

```


```{r}
#| echo: false

top_10 <- measles_2024 |> arrange(desc(measles_incidence_rate_per_1000000_total_population)) |>
  slice(1:10) |>
  mutate(country = fct_reorder(country,
    measles_incidence_rate_per_1000000_total_population))

top10_inc <- ggplot(data = top_10, aes(x = country, y = measles_incidence_rate_per_1000000_total_population, text = paste0("Country: ", country, "\n",
    "Incidence: ", measles_incidence_rate_per_1000000_total_population))) +
  geom_col(aes(fill = region)) +
  theme_minimal() +
  labs(y = "Incidence rate per million") +
  coord_flip() +
  scale_fill_viridis_d()

ggplotly(top10_inc, tooltip = "text")

```

From the choropleth map and the barplot of the top ten countries with the highest incidence rate above, we can see that incidence rates per million of the population tell a different story from that told by total reported cases. Kyrgyzstan is now at the top with the highest incidence rate (2005 cases per million). This means that in 2024, people in Kyrgyzstan were at the highest risk of being infected with measles as compared to other countries.

The EURO region still has the highest number of countries even in the top ten countries with the highest incidence rates.

The bar chart confirms the patterns observed in the choropleth map, with countries such as Kazakhstan, Kyrgyzstan, Azerbaijan, and Romania exhibiting the highest incidence rates. This observation led me to discover that there was in fact a measles outbreak in the EURO region (European and Central Asian Region) in 2024 as reported on a press release by UNICEF.


# Comparing Lesotho and USA

Moving on from a global analysis of measles cases and incidence rates, as an attempt to answer my question about why I went from hearing a lot about measles when I was in Lesotho to not hearing about it all here in the United States, in this section I take a close-up look at the differences in measles reports between these two countries.

Reports for the U.S. only begin in 2014 while reports for other countries in this data set start from 2012. This is likely due to the fact that USA had been considered to have eliminated measles since 2000 because it had been having incidence rates of less that 1 per million. So what happened in 2014 to reverse this elimination? 

After doing a bit of searching, I found several sources pointing this resurgence of measles to Disneyland! One of these sources is an article titled "Measles at Disneyland, a Problem for All Ages", published by Dr. Halsey and Dr. Salmon in 2015. This article reports an outbreak that affected Disneyland guests and employees, which was suspected to have been sparked by an international traveler. The rapid spread of measles among people of all ages in this outbreak was assosiated with a lack of immunization as measles had no longer been a threat for over a decade until then. 

```{r}
#| echo: false
US_Les <- measles |> filter(country == "Lesotho"| country == "United States of America")

ggplot(data = US_Les, aes(x = year, y = measles_total)) +
  geom_line(aes(colour = country), linewidth = 1.2) +
  scale_color_viridis_d() +
  scale_x_continuous(breaks = seq(2012, 2024, by = 2)) +
  scale_y_continuous(limits = c(0,NA)) +
  theme_minimal() +
  labs(x = "Year",
       y = "Total cases reported")
```


```{r}
peak_Les <- US_Les |> 
  filter(country == "Lesotho") |> 
  arrange(desc(measles_incidence_rate_per_1000000_total_population)) |>
  slice(1)
 

ggplot(data = US_Les, aes(x = year, y = measles_incidence_rate_per_1000000_total_population)) +
  geom_line(aes(colour = country), linewidth = 1.2) +
  geom_label(data = peak_Les, 
            aes(label =measles_incidence_rate_per_1000000_total_population)) +
  scale_color_viridis_d() +
  scale_x_continuous(breaks = seq(2012, 2024, by = 2)) +
  scale_y_continuous(limits = c(0,NA)) +
  theme_minimal() +
  labs(x = "Year",
       y = "Incidence rate per million") 

```

Looking at the two line plots above, where the first shows trends in total measles cases reported for Lesotho and USA and the second shows trends in incidence rates in these countries, they bring into perspective how different a story total case count tells from incidence rate. On the first plot, we can clearly see that despite their fluctuation through time, the total reported cases in USA are constantly higher than reported cases in Lesotho.  

However, on the second plot, we can see that the incidence rate in Lesotho tends to fluctuate to levels higher than the U.S.

Even though the total cases for Lesotho seem relatively steady in comparison to those of the U.S, the incidence rate in Lesotho is shown to be more volatile in the second plot.This goes to demonstrate the effect of incidence rate in leveling the playing field for countries of different population sizes when comparing the risk of a disease. From the second plot, it can be concluded that risk of measles is higher in Lesotho than in the U.S.

The several peaks in incidence rates for Lesotho most likely coincide with measles outbreaks. The most interesting and extreme one is the peak of 2013 which goes as high as approximately 72 cases per million of the population. This was during a region-wide outbreak across Southern Africa which was accelerated by low immunization as vaccination coverage had dropped in that period.





# even though incidence rate is a better indicator of prevalence than total cases because it accounts for the differences in population, it also raises some limitations:

# it overly magnifies prevalence in smaller countries even through their counts or cases may be lower  -- therefore requires a well-rounded perspective when comparing countries of different population sizes

# 330–340 million people USA ,,,, 2.1–2.3 million people Lesotho


## One might also suggest that perhaps the global trend in population count may give some insight into the trends in measles cases. Wouldn't it make sense that assuming all things are held constant, as population grows the number of rate of infection will grow thus increasing the incidence rate as well as the total number of reported cases? In fact, I suspect that /.....



USA and Lesotho a greatly different economically...

## this then begs the question, does economic status/economic level have any influence on incidence rates -- year, GDP, population size, facet by region?
-- predicting the incidence rates in USA AND LES in 2027??

1. How have measles cases changed globally over time?

# Visualization:

Line plot (year vs. measles incidence rate, grouped by region or global average)

```{r}
View(measles)
```


```{r}
#Line plot for global trend in total cases of measles over the years

annual_measles <- measles |> group_by(year) |>
  summarise(`total measles cases in millions` = sum(measles_total)/100000)
  
ggplot(data = annual_measles, aes(x = year, y = `total measles cases in millions`)) +
  geom_line(color = "darkblue") +
  theme_minimal() +
  labs(title = "Trend of measles cases throughout the world for the past 13 years",
       x = "Years",
       y = "Total global measles cases in millions")

# the sharp spike between from 2018 to 2019??

# dropped significantly from 2019 to 2021 because people were isolating in those years due to the Covid pandemic. Since measles is a contagious viral infection, its spread got minimized by the fact that people did not come into contact with other people as frequently as they used to.

#The sharp decline in measles incidence between 2019 and 2021 may be partially explained by reduced social contact during the COVID-19 pandemic. However, it may also reflect disruptions in surveillance and reporting systems, meaning the true number of cases could be underreported during this period.

#how does this trend compare with the trend in global population growth?
annual_population <- measles |> group_by(year) |>
  summarise(`total world population in billions` = sum(total_population)/1000000000)

ggplot(data = annual_population, aes(x = year, y = `total world population in billions`)) +
  geom_line(color = "skyblue") +
  theme_minimal() +
  labs(title = "Trend of measles cases throughout the world for the past 13 years",
       x = "Years",
       y = "Total world population in billions")

## we see a generally upward trend in global population 
## According to the Population Reference Bureau, the world population in 2012 was 7.1 billion (https://www.prb.org/resource/2012-world-population-data-sheet/), so this suggests that there was error in my computation of total world population for the years 2012 and 2013 which seem to be significantly lower on the plot. Looking at the measles_month data set, the most likely cause of this discrepancy is the missing This discrepancy causes a limitation to the analysis, but even despite it, there still has been an overall growth in global population with some declines between 2020 and 2021 due to the Covid-19 pandemic which increased global death rates. 

# global measles trends don't seem to be linked to trends in global population growth.
```

```{r}
# trends over time for different regions

```

2. How does the geographic distribution of measles cases vary across countries and regions?


#Visualization:

Choropleth world map (fill = incidence rate, possibly animated over time)

```{r}
# we have seen the trends in global measles cases over the years, now to explore the distribution of measles cases across the world, we will look at a choropleth world map using the most recent year with complete records - 2024.

## we will use incidence rate per million

measles_2024 <- measles |> filter(year == 2024)

library(maps)
world <- map_data("world")

measles_2024 <- measles_2024 |> mutate(country = if_else(country == "United States of America",
                                   true = "USA",
                                   false = country))

measles_2024 <- measles_2024 |> mutate(country = if_else(country == "Russian Federation",
                                   true = "Russia",
                                   false = country))
measles_2024 <- measles_2024 |> mutate(country = if_else(country == "Eswatini",
                                   true = "Swaziland",
                                   false = country))

world_measles <- measles_2024 |> left_join(world, by = join_by(country == region))

plot1 <- ggplot(data = world_measles, aes(x = long, y = lat, group = group, text = paste0(
    "Country: ", country, "\n",
    "Incidence: ", round(measles_incidence_rate_per_1000000_total_population, 2)
  ))) +
  geom_polygon(colour = "black", aes(fill = measles_incidence_rate_per_1000000_total_population)) +
  coord_map(projection = "gilbert") +
  theme_void() +
  scale_fill_distiller(type = "seq", palette = "Blues", direction = 1) +
  xlim(c(-180, 180)) +
  labs(fill = "Incidence rate per million")


plot <- ggplot(world_measles, aes(
  x = long,
  y = lat,
  group = group,
  text = paste0(
    "Country: ", country, "\n",
    "Incidence: ", round(measles_incidence_rate_per_1000000_total_population, 2)
  )
)) +
  geom_polygon(
    aes(fill = measles_incidence_rate_per_1000000_total_population),
    color = "gray40",
    size = 0.1
  ) +
  coord_map("gilbert") +
  theme_void() +
  scale_fill_distiller(
    palette = "YlOrRd",
    direction = 1,
    na.value = "gray90"
  ) +
  labs(fill = "Incidence per 1,000,000")

library(plotly)

ggplotly(plot1, tooltip = "text")


# highest incidence rates in Kazakhstan, Kyrgyzstan, Azerbaijan and Romania
```


3. Which countries and regions experience the highest measles burden, and how has this changed over time?

3.  Which countries and regions experienced the highest measles incidence rates in 2024?

#Visualization:

Bar chart (top countries by incidence or total cases)
Optional: small multiples or faceted trends by region

```{r}
# does the choropleth map display accurate results fo the distribution of measles incidence rates across the world?

# to check this, below is a bar chart of the countries with the highest incidence rates in 2024
library(forcats)

top_10 <- measles_2024 |> arrange(desc(measles_incidence_rate_per_1000000_total_population)) |>
  slice(1:10) |>
  mutate(country = fct_reorder(country,
    measles_incidence_rate_per_1000000_total_population))

View(top_10)

ggplot(data = top_10, aes(x = country, y = measles_incidence_rate_per_1000000_total_population)) +
  geom_col(aes(fill = region)) +
  theme_minimal() +
  labs(y = "Incidence rate per million") +
  coord_flip()

# The bar chart confirms the patterns observed in the choropleth map, with countries such as Kazakhstan, Kyrgyzstan, Azerbaijan, and Romania exhibiting the highest incidence rates. This reinforces the observation that high measles burden was not confined to a single region in 2024, but was higher in contries in both Central Asia and Eastern Europe.

## https://www.unicef.org/kyrgyzstan/press-releases/european-and-central-asian-region-reports-highest-number-measles-cases-more-25-years

# European and central asian outbreak
```




- questions of interest to be explored in the project

[does GDP have anything to do with it -- economic disparities] - as a student of Economics, I cannot help but wonder what role - if any - economic factors such as GDP per capita play in the disparities in measles incidence across the world


# comparing Lesotho with USA

The US is said to have eliminated measles since 2000. But why the spike in 2019? and why not 0 incidence rates throughout?

```{r}
# line plot
US_Les <- measles |> filter(country == "Lesotho"| country == "United States of America") |>
       filter(year >= 2014)

# USA data only starts from 2014, hence the filtering

ggplot(data = US_Les, aes(x = year, y = measles_incidence_rate_per_1000000_total_population)) +
  geom_line(aes(colour = country)) +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(x = "Year",
       y = "Incidence rate per million")

# Why the spike in 2019 for both countries? 
# why the 2 other spikes in 2016 and 2023 for Lesotho? -- likely due to fluctuations in reporting where the "dips" may just be due to underreporting. Could also be due to small ourbreaks seeming larger because of the smaller population compared to the US.
# - this shows how measles reports (incidence rates) in smaller countries can appear more extreme even when driven by relatively similar case counts.

## The comparison between Lesotho and United States highlights substantial differences in both the level and variability of measles incidence. Lesotho exhibits more frequent spikes, which may reflect localized outbreaks or differences in vaccination coverage. In contrast, the United States shows generally lower and more stable incidence, likely due to stronger vaccination programs and surveillance systems.

# comparing reporting for both countries, 
# interpreting what this means in relation to their widely different population sizes
# comparing with economic factors and linking this to reporting and confirmation of cases
# stacked bar plot to compare proportions of confirmed, clinical and epi-linked cases --
# anything to do with economy and medical advancements like labs?
```

## heatmap for seasonality for one country -- to see if the disease is seasonal


# Findings

- the most interesting visualizations + their interpretations  --- broken up according to the questions of interest being answered

- explain why results form these visualizations matter

- state potential flaws to the visualizations and the interpretations



# Conclusion

- future work that could be done if there was more time

- limitations to my visualizations

* the interactive map (with leaflet) -- limitation is that it depends on the user hovering over it to see the exact incidence rates, which is a limitaTion of interactive visualisations --- ADD MORE TO THIS 

## Modeling mealses cases using predictors that are economic, and maybe rubella cases as well?/??

Future work:

vaccination data
policy analysis

## LIMITATIONS

some countries had inconsistent data reporting eg in 2024, Botswana had monthly data for Jan to Sep then skipped Oct and Nov and had Dec


Filtered out data for the year 2025 because it was mostly incomplete, only had the first 5 or 6 months' data, so for the sake of consistency, I removed it.

Underreporting in some countries

countries recorded with different names in the 2 data sets for world and measles. made it harder to merge the 2 accurately, had to go in manually and change some names to match. I may have missed some countries
Like: 

filter(country == "Eswatini")

world |> filter(region == "Swaziland")


# References 

https://stackoverflow.com/questions/13701347/force-the-origin-to-start-at-0 

https://stackoverflow.com/questions/71628944/custom-tooltip-hovertext-for-choropleth-created-with-plotly-via-r

https://www.worldometers.info/world-population/iraq-population/#:~:text=Country%20Continent%20World,in%20Iraq%20is%2021.1%20years. 

https://www.worldometers.info/world-population/namibia-population/

https://public-health.tamu.edu/degrees/mph/blog/epidemiology-incidence-vs-prevalence-explained.html

https://www.unicef.org/kyrgyzstan/press-releases/european-and-central-asian-region-reports-highest-number-measles-cases-more-25-years

https://www.cdc.gov/mmwr/preview/mmwrhtml/mm5106a2.htm

https://www.acpjournals.org/doi/10.7326/M15-0447 

https://reliefweb.int/report/lesotho/southern-africa-measles-back-vengeance

